### 2024-05-22 thesis/input_files/paper1.tex
22 May 2024

We take then the inner product in the high dimensional space $\mathcal{H}$ mapped by $\phi(x)$, i.e.\
\\
\begin{align*}
k(x,y):=&\langle \phi(x), \phi(y) \rangle_{\mathcal{H}}
\end{align*}
\\
where $\phi(x)$ is referred to as feature map while $k$ is the kernel function.
\\
Therefore, we can kernelize any algorithm involving a dot product by substituting $\langle x, y \rangle_{\mathcal{X}}$ with $\langle \phi(x), \phi(y) \rangle_{\mathcal{H}}$
\\
One would expect constructing the feature maps explicitly and then evaluating their inner product in $\mathcal{H}$ to be computationally expensive, and indeed it is. However, we do not have to explicitly perform such calculations. This is because of the kernel trick.
\\
To illustrate the idea behind the kernel trick consider the following example.
Suppose $x, y \in \mathbb{R}^2$ and assume $\phi(x)=(x_{1}^{2},x_{2}^{2},\sqrt{2}x_{1}x_{2})$, then the inner product in the feature space is $x_{1}^{2}y_{1}^{2},x_{2}^{2}y_{2}^{2}+2x_{1}x_{2}y_{1}y_{2}$.
Notice that this is the same of $\langle \phi(x), \phi(y) \rangle_{\mathcal{H}}$; thus the kernel trick consists of just using $k(x,y):=(x^\intercal y)^2$.
For instance:
\\
$\langle \phi([2,4]),\phi([8,9]) \rangle_{\mathcal{H}}=  \left[ {\begin{array}{cccc}
    4 & 16 &     \sqrt{2}\cdot 2 \cdot 4\\
  \end{array} } \right]
  \left[ {\begin{array}{cccc}
    64\\
    81\\
    \sqrt{2}\cdot8 \cdot 9
    \\
  \end{array} } \right]\\=4\cdot 64 +16\cdot 81 +2 \cdot 2 \cdot 4 \cdot 8 \cdot 9=2704
$
\\
\\
While $k([2,4], [8,9])=(16+36)^2=2704$
We can see that the latter calculation is much quicker and compact.
This gives a taste of how powerful kernels are thanks to the kernel trick.
Indeed, the idea of the the kernel trick can be extended to feature maps
$\phi$ involving an infinite feature space. In these cases, calculating the
respective kernel is equal to calculating the inner product between an infinite
number of features of the data points.

@@

Next, we make this characterisation explicit by showing how to construct a feature mapping $\phi$ of a kernel $k$ satisfying the finitely positive semidefinite property.
Let the feature space be the set of functions 
\\
$\mathcal{F} =\{ \sum\limits_{i=1}^n \alpha_i k(x_i, \cdot): n \in \mathbb{N}, x_i, y_i \in \mathcal{X}, a_i \in \mathbb{R}, i=1,\cdots, n\}$.
This space is clearly closed under addition of functions and scalar multiplication, therefore it is a vector space.
Letting $f(x)=\sum\limits_{i=1}^n \alpha_i k_i(x_i,x)$ and $g(x)=\sum\limits_{i=1}^m \beta_i k_i(y_i,x)$,  we can introduce an inner product on $\mathcal{F}$ as 
\\
$\langle f,g \rangle_{\mathcal{F}}=:\sum\limits_{i=1}^n\sum\limits_{j=1}^m a_i \beta_j k_i(x_i,y_j)=\sum\limits_{i=1}^n a_i g(x_i)= \sum\limits_{j=1}^m \beta_j f(y_j)$
\\
Notice, $\langle f, f \rangle_{\mathcal{F}}= \alpha^\intercal K \alpha \geq 0 \ \forall f \in \mathcal{F}$, where $K$ is the kernel matrix.
Moreover, $\langle f, g \rangle_{\mathcal{F}}$ is real valued, symmetric and bilinear. All these facts implies that the just introduced inner product is a valid metric.
\\
Additionally, letting $g(x)=k(\cdot, x)$, we have that 
\begin{equation}
    \langle f, k(\cdot, x)\rangle_{\mathcal{F}}=\sum\limits_{i=1}^n a_i k(x_i,x)=f(x)    
\end{equation}
This is known as the reproducing property.
\\
Finally, to show the characterisation of kernels we have to check that the inner product makes $\mathcal{F}$ a complete metric space.
Consider a fixed input $x$ and a Cauchy sequence $\{f_n\}_{n=1}^{\infty}$, we then have by means of the Cauchy-Schwarz inequality that 
\\
\begin{equation}
    (f_n(x)- f_m(x))^2=\langle f_n-f_m, k(\cdot, x)\rangle_{\mathcal{F}}^2 \leq \| f_n - f_m \|_{\mathcal{F}}^2 k(x,x)    
\end{equation}
\\
We have then, that $f_n(x)$ is a Cauchy sequence of real numbers. Thus, we can let $g(x)=\lim_{n\to \infty} f_n(x)$ and include all such limits functions in $\mathcal{F}$, doing so we complete $\mathcal{F}$ and we obtain the Hilbert space $\mathcal{H}$ associated to the kernel $k$.
Having constructed the feature space $\mathcal{H}$, the mapping $\phi(x)$ is defined as follows
\\
\begin{equation}
    \phi: x \in \mathcal{X} \to \phi(x)=k(\cdot, x) \in \mathcal{H}
\end{equation}
\\
Notice, that by the reproducing property, $f$ can be represented as the inner product with itself in the feature space, that is 
\begin{equation}
    f(x)=\langle f, \phi(x)\rangle_{\mathcal{H}}=\langle f, k(\cdot, x)\rangle_{\mathcal{H}}=f(x)
\end{equation}

\begin{definition}
    Let $(\mathcal{H}, \langle \cdot, \cdot \rangle_\mathcal{H})$ be a Hilbert space of real-valued functions on $\mathcal{X}$. A function $k: \mathcal{X} \times \mathcal{X} \to \mathbb{R}$ is called a reproducing kernel of $\mathcal{H}$ if and only if 
     \begin{align}
        k(x, \cdot) \in \mathcal{H} &\quad \ \forall x \in \mathcal{X}    \\
        \langle  f, k(x, \cdot) \rangle_\mathcal{H} = f(x) &\quad \ \forall f\in \mathcal{H}, x \in \mathcal{X}
    \end{align}
    Notice, the finitely positive semidefinite property of kernel $k$ is a sufficient condition for $\mathcal{H}$ to have an associated RKHS.
\end{definition}

@@

The function $k(\cdot, x)$ is the reproducing kernel evaluated at point $x$.
Furthermore, note that $k_{x}$ is itself a function lying in it.

@@

\begin{proof}
    Consider a finite set of points $x_1, \cdots, x_n$, a vector $z \in \mathbb{R}^n$ and let $K_1$ and $K_2$ be the kernel matrices associated to $k_1$ and $k_2$ evaluted on these points.
    \\
    We have:
    \\
    \begin{align}
    z^\intercal (K_1+K_2) z= z^\intercal K_1 z + z^\intercal K_2 z \geq 0
    \\
    z^\intercal aK z= a z^\intercal K z \geq 0
    \end{align}
    Let $K=K_1 \otimes K_2$, we have that the eigenvalues of $K$ are made up by the product pairs of the eigenvalues of $K_1$ and $K_2$. Thus, $K$ is positive semidefinite. Next, the matrix given by $k_1 \cdot k_2$ corresponds to the Schur product of $K_1$ and $K_2$, denote it $H$. Notice, $H$ is a principal submatrix of $K$, hence for any $z \in \mathbb{R}^n$ there exist a $t \in \mathbb{R}^{n^2}$ such that $z^\intercal H z= t^\intercal K t$. Using the fact that K is positive semidefinite we have that for any $z$, it holds 
    \begin{align}
        z^\intercal H z= t^\intercal K t\geq 0
    \end{align}
\end{proof}

@@

Notice, in the Matern family $\Gamma(\cdot)$ is the gamma function while $Y(\cdot)$ is the Bessel function of the second kind.
\\
When $\nu=p+\frac{1}{2}$, $p \in \mathbb{N}^+$, the Matern kernel can be rewritten as a product of an exponential and a polynomial of degree $p$. 
\\
$k(x,y)=\exp \left(-{\frac {{\sqrt {2p+1}}\|x-y|\|_{2}}{l }}\right){\frac {p!}{(2p)!}}\sum\limits_{i=0}^{p}{\frac {(p+i)!}{i!(p-i)!}}\left({\frac {2{\sqrt {2p+1}}\|x-y|\|_{2}}{l }}\right)^{p-i}$
\\
For example:
\begin{itemize}
    \item $p=0,\nu=\frac{1}{2} \implies k(x,y)=\exp\left(\frac{-\|x-y|\|_{2}}{l}\right)$
    \item $p=1,\nu=\frac{3}{2} \implies 
    k(x,y)=\left(1+{\frac {{\sqrt {3}}\|x-y|\|_{2}}{l }}\right)\exp \left(-{\frac {{\sqrt {3}}\|x-y|\|_{2}}{l }}\right)$
    \item $p=2,\nu=\frac{5}{2} \implies k(x,y)=\left(1+{\frac {{\sqrt {5}}\|x-y|\|_{2}}{l }}+{\frac {5\|x-y|\|_{2}^{2}}{3 l ^{2}}}\right)\exp \left(-{\frac {{\sqrt {5}}\|x-y|\|_{2}}{l }}\right)$
\end{itemize}

>>>>

We take then the inner product in the high dimensional space $\mathcal{H}$ mapped by $\phi(x)$, i.e.\
\\
\begin{align*}
k(x,y):=&\langle \phi(x), \phi(y) \rangle_{\mathcal{H}}
\end{align*}
\\
where $\phi(x)$ is referred to as feature map while $k$ is the kernel function.
\\
Therefore, we can kernelize any algorithm involving a dot product by substituting $\langle x, y \rangle_{\mathcal{X}}$ with $\langle \phi(x), \phi(y) \rangle_{\mathcal{H}}$
\\
One would expect constructing the feature maps explicitly and then evaluating their inner product in $\mathcal{H}$ to be computationally expensive, and indeed it is. However, we do not have to explicitly perform such calculations. This is because of the kernel trick.
\\
To illustrate the idea behind the kernel trick consider the following example.
Suppose $x, y \in \mathbb{R}^2$ and assume $\phi(x)=(x_{1}^{2},x_{2}^{2},\sqrt{2}x_{1}x_{2})$, then the inner product in the feature space is $x_{1}^{2}y_{1}^{2}+x_{2}^{2}y_{2}^{2}+2x_{1}x_{2}y_{1}y_{2}$.
Notice that this is the same of $\langle \phi(x), \phi(y) \rangle_{\mathcal{H}}$; thus the kernel trick consists of just using $k(x,y):=(x^\intercal y)^2$.
For instance:
\\
$\langle \phi([2,4]),\phi([8,9]) \rangle_{\mathcal{H}}=  \left[ {\begin{array}{cccc}
    4 & 16 &     \sqrt{2}\cdot 2 \cdot 4\\
  \end{array} } \right]
  \left[ {\begin{array}{cccc}
    64\\
    81\\
    \sqrt{2}\cdot8 \cdot 9
    \\
  \end{array} } \right]\\=4\cdot 64 +16\cdot 81 +2 \cdot 2 \cdot 4 \cdot 8 \cdot 9=2704
$
\\
\\
While $k([2,4], [8,9])=(16+36)^2=2704$
We can see that the latter calculation is much quicker and compact.
This gives a taste of how powerful kernels are thanks to the kernel trick.
Indeed, the idea of the the kernel trick can be extended to feature maps
$\phi$ involving an infinite feature space. In these cases, calculating the
respective kernel is equal to calculating the inner product between an infinite
number of features of the data points.

@@

Next, we make this characterisation explicit by showing how to construct a feature mapping $\phi$ of a kernel $k$ satisfying the finitely positive semidefinite property.
Let the feature space be the set of functions 
\\
$\mathcal{F} =\{ \sum\limits_{i=1}^n \alpha_i k(x_i, \cdot): n \in \mathbb{N}, x_i, y_i \in \mathcal{X}, a_i \in \mathbb{R}, i=1,\cdots, n\}$.
This space is clearly closed under addition of functions and scalar multiplication, therefore it is a vector space.
Letting $f(x)=\sum\limits_{i=1}^n \alpha_i k_i(x_i,x)$ and $g(x)=\sum\limits_{i=1}^m \beta_i k_i(y_i,x)$,  we can introduce an inner product on $\mathcal{F}$ as 
\\
$\langle f,g \rangle_{\mathcal{F}}:=\sum\limits_{i=1}^n\sum\limits_{j=1}^m a_i \beta_j k_i(x_i,y_j)=\sum\limits_{i=1}^n a_i g(x_i)= \sum\limits_{j=1}^m \beta_j f(y_j)$
\\
Notice, $\langle f, f \rangle_{\mathcal{F}}= \alpha^\intercal K \alpha \geq 0 \ \forall f \in \mathcal{F}$, where $K$ is the kernel matrix.
Moreover, $\langle f, g \rangle_{\mathcal{F}}$ is real valued, symmetric and bilinear. All these facts imply that the just introduced inner product is a valid metric.
\\
Additionally, letting $g(x)=k(\cdot, x)$, we have that 
\begin{equation}
    \langle f, k(\cdot, x)\rangle_{\mathcal{F}}=\sum\limits_{i=1}^n a_i k(x_i,x)=f(x)    
\end{equation}
This is known as the reproducing property.
\\
Finally, to show the characterisation of kernels we have to check that the inner product makes $\mathcal{F}$ a complete metric space.
Consider a fixed input $x$ and a Cauchy sequence $\{f_n\}_{n=1}^{\infty}$, we then have by means of the Cauchy-Schwarz inequality that 
\\
\begin{equation}
    (f_n(x)- f_m(x))^2=\langle f_n-f_m, k(\cdot, x)\rangle_{\mathcal{F}}^2 \leq \| f_n - f_m \|_{\mathcal{F}}^2 k(x,x)    
\end{equation}
\\
We have then, that $f_n(x)$ is a Cauchy sequence of real numbers. Thus, we can let $g(x):=\underset{n\to \infty}\lim f_n(x)$ and include all such limits functions in $\mathcal{F}$, doing so we complete $\mathcal{F}$ and we obtain the Hilbert space $\mathcal{H}$ associated to the kernel $k$.
Having constructed the feature space $\mathcal{H}$, the mapping $\phi(x)$ is defined as follows
\\
\begin{equation}
    \phi: x \in \mathcal{X} \to \phi(x)=k(\cdot, x) \in \mathcal{H}
\end{equation}
\\
Notice, that by the reproducing property, $f$ can be represented as the inner product with itself in the feature space, that is 
\begin{equation}
    f(x)=\langle f, \phi(x)\rangle_{\mathcal{H}}=\langle f, k(\cdot, x)\rangle_{\mathcal{H}}=f(x)
\end{equation}

\begin{definition}
    Let $(\mathcal{H}, \langle \cdot, \cdot \rangle_\mathcal{H})$ be a Hilbert space of real-valued functions on $\mathcal{X}$. A function $k: \mathcal{X} \times \mathcal{X} \to \mathbb{R}$ is called a reproducing kernel of $\mathcal{H}$ if and only if 
     \begin{align}
        k(x, \cdot) \in \mathcal{H} &\quad \ \forall x \in \mathcal{X}    \\
        \langle  f, k(x, \cdot) \rangle_\mathcal{H} = f(x) &\quad \ \forall f\in \mathcal{H}, \ x \in \mathcal{X}
    \end{align}
    Notice, the finitely positive semidefinite property of kernel $k$ is a sufficient condition for $\mathcal{H}$ to have an associated RKHS.
\end{definition}

@@

The function $k(\cdot, x)$ is the reproducing kernel evaluated at point $x$.
Furthermore, note that $k(\cdot, x)$ is itself a function lying in it.

@@

\begin{proof}
    Consider a finite set of points $x_1, \cdots, x_n$, a vector $z \in \mathbb{R}^n$ and let $K_1$ and $K_2$ be the kernel matrices associated to $k_1$ and $k_2$ evaluated on these points.
    \\
    We have:
    \\
    \begin{align}
    z^\intercal (K_1+K_2) z&= z^\intercal K_1 z + z^\intercal K_2 z \geq 0
    \\
    z^\intercal aK z&= a z^\intercal K z \geq 0
    \end{align}
    Let $K=K_1 \otimes K_2$, we have that the eigenvalues of $K$ are made up by the product pairs of the eigenvalues of $K_1$ and $K_2$. Thus, $K$ is positive semidefinite. Next, the matrix given by $k_1 \cdot k_2$ corresponds to the Schur product of $K_1$ and $K_2$, denote it $H$. Notice, $H$ is a principal submatrix of $K$, hence for any $z \in \mathbb{R}^n$ there exist a $t \in \mathbb{R}^{n^2}$ such that $z^\intercal H z= t^\intercal K t$. Using the fact that K is positive semidefinite we have that for any $z$, it holds 
    \begin{align}
        z^\intercal H z= t^\intercal K t\geq 0
    \end{align}
\end{proof}

@@

Notice, in the Matern family, $\Gamma(\cdot)$ is the gamma function while $Y(\cdot)$ is the Bessel function of the second kind.
\\
When $\nu=p+\frac{1}{2}$, $p \in \mathbb{N}^+$, the Matern kernel can be rewritten as a product of an exponential and a polynomial of degree $p$. 
\\
$k(x,y)=\exp \left(-{\frac {{\sqrt {2p+1}}\|x-y|\|_{2}}{l }}\right){\frac {p!}{(2p)!}}\sum\limits_{i=0}^{p}{\frac {(p+i)!}{i!(p-i)!}}\left({\frac {2{\sqrt {2p+1}}\|x-y|\|_{2}}{l }}\right)^{p-i}$
\\
For example:
\begin{itemize}
    \item $p=0,\nu=\frac{1}{2} \implies k(x,y)=\exp\left(-\frac{\|x-y|\|_{2}}{l}\right)$
    \item $p=1,\nu=\frac{3}{2} \implies 
    k(x,y)=\left(1+{\frac {{\sqrt {3}}\|x-y|\|_{2}}{l }}\right)\exp \left(-{\frac {{\sqrt {3}}\|x-y|\|_{2}}{l }}\right)$
    \item $p=2,\nu=\frac{5}{2} \implies k(x,y)=\left(1+{\frac {{\sqrt {5}}\|x-y|\|_{2}}{l }}+{\frac {5\|x-y|\|_{2}^{2}}{3 l ^{2}}}\right)\exp \left(-{\frac {{\sqrt {5}}\|x-y|\|_{2}}{l }}\right)$
\end{itemize}

### 2024-05-22 thesis/input_files/problemdescription.tex
22 May 2024

Individuals and organizations constantly face situations of uncertainty, thus the need for robust forecasting methods. Such methods are crucial in the process of taking informed decisions and for strategic planning.
\\
The basic idea of forecasting is that we can extract knowledge from the past in order to make educated guesses about the future. Consequently, the range of fields where forecasting can be applied is very wide.
In this thesis, our focus lies on applying forecasting techniques to the energy sector. 
\\
Our decision to focus on the energy market is mainly motivated by the rapid changes it has undergone. %/experienced 
Over the last decades, electricity markets have gone through an unprecedented transformation. This shift was driven by the liberalization of such markets, the development and integration of renewable energy sources, the  increase of low carbon technologies and the adoption of smart meters. Events like the California electricity crisis are further motivating the choice of the electricity sector as subject of our studies, see \cite{california}.
Moreover, the process of deregulation lead to an increasing interest in the field of electricity forecasting (EF) within the academic community, see figure \ref{fig:epf_evolution}.
In addtion, the United Nations have identified the right to access affordable, reliable, sustainable and modern energy as one of their 17 sustainable development goals (SDGs) \cite{un_sdgs}.
Finally, the electricity market has a set of features that make it unique: electricity cannot be stored in an efficient way and supply and demand have to be matched instantly.
\\
\section{Motivation}
There are mutliple reasons why the energy sector needs robust forecasting techniques.
For power market companies, being able to predict prices with a low mean absolute percentage error (MAPE) \ref{mape} results in increased savings \cite{savings}. Furthermore, the adoption of smart meters provides power market companies with a huge amount of consumer data. This enables them to better model consumer preferences.
\\
Transmission system operators' (TSO) main goal is to match supply and demand, generally TSO do so by increasing or decreasing the generation. Thus, from their point of view forecasting is critical for balancing the electricity network.
Probabilistic forecasting may be useful to power producers, traders and consumers in order to improve their decision making process and managing risk. This holds in particular for traders, because probabilistic forecasts enable them to simulate scenarios and carry out stress tests.
Other possible applications include: control of storage, demand side response, anomaly detection, network design and planning, simulating inputs and handling missing data.
\\
\section{Point versus probabilistic forecast}
A distinction has to be made between two types of forecasting approaches: point forecasts and probabilistic forecasts.
Point prediction, also called deterministic forecasting in the literature \cite{EPF_review}, is all about predicting a particular value in time.
On the other hand, with probabilistic forecasting we aim at predicting either an interval, quantiles or a probability distribution for each point in time \cite{nowotarski}. For this reason, probabilistic forecasts are more informative than point forecasts. This is why the interest of the research community is shifting towards them.
A probabilistic forecast can be turned into a point forecast by simply taking its expectation.
Alternatively, a probabilistic forecast can be derived from a point one by modeling the residuals of the point prediction.

>>>>

Individuals and organizations constantly face situations of uncertainty, thus the need for robust forecasting methods. Such methods are crucial in the process of taking informed decisions and for strategic planning.
\\
The basic idea of forecasting is that we can extract knowledge from the past in order to make educated guesses about the future. Consequently, the range of fields where forecasting can be applied is very wide.
In this thesis, our focus lies on applying forecasting techniques to the energy sector. 
\\
Our decision to focus on the energy market is mainly motivated by the rapid changes it has undergone. %/experienced 
Over the last decades, electricity markets have gone through an unprecedented transformation. This shift was driven by the liberalization of such markets, the development and integration of renewable energy sources, the  increase of low carbon technologies and the adoption of smart meters. Events like the California electricity crisis are further motivating the choice of the electricity sector as subject of our studies, see \cite{california}.
Moreover, the process of deregulation lead to an increasing interest in the field of electricity forecasting (EF) within the academic community, see figure \ref{fig:epf_evolution}.
In addtion, the United Nations have identified the right to access affordable, reliable, sustainable and modern energy as one of their 17 sustainable development goals (SDGs) \citeW{un_sdgs}.
Finally, the electricity market has a set of features that make it unique: electricity cannot be stored in an efficient way and supply and demand have to be matched instantly.
\\
\section{Motivation}
There are mutliple reasons why the energy sector needs robust forecasting techniques.
For power market companies, being able to predict prices with a low mean absolute percentage error (MAPE) \ref{mape} results in increased savings \cite{savings}. Furthermore, the adoption of smart meters provides power market companies with a huge amount of consumer data. This enables them to better model consumer preferences.
\\
Transmission system operators' (TSO) main goal is to match supply and demand, generally TSO do so by increasing or decreasing the generation. Thus, from their point of view forecasting is critical for balancing the electricity network.
Probabilistic forecasting may be useful to power producers, traders and consumers in order to improve their decision making process and managing risk. This holds in particular for traders, because probabilistic forecasts enable them to simulate scenarios and carry out stress tests.
Other possible applications include: control of storage, demand side response, anomaly detection, network design and planning, simulating inputs and handling missing data.
\\
\section{Point versus probabilistic forecast}
A distinction has to be made between two types of forecasting approaches: point forecasts and probabilistic forecasts.
Point prediction, also called deterministic forecasting in the literature \cite{EPF_review}, is all about predicting a particular value in time.
On the other hand, with probabilistic forecasting we aim at predicting either an interval, quantiles or a probability distribution for each point in time \cite{nowotarski}. For this reason, probabilistic forecasts are more informative than point forecasts. This is why the interest of the research community is shifting towards them.
A probabilistic forecast can be turned into a point forecast by simply taking its expectation.
Alternatively, a probabilistic forecast can be derived from a point one by modeling the residuals of the point prediction.
